---
title: "Paper Outline: The Global Sparsity and Regulatory Rigidity of Fecal Indicator Bacteria Monitoring"
author: "Matthew R. V. Ross"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(arrow)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(sf)
library(rnaturalearth)
library(patchwork)

# Helper function to insert explicit NAs to break lines over winter/off-season gaps
insert_gaps <- function(df, max_gap) {
  if (nrow(df) < 2) return(df)
  df <- df[order(df$date), ]
  new_rows <- list()
  for (i in 1:(nrow(df) - 1)) {
    new_rows[[length(new_rows) + 1]] <- df[i, ]
    gap <- as.numeric(df$date[i+1] - df$date[i])
    if (gap > max_gap) {
      na_row <- df[i, ]
      na_row$date <- df$date[i] + 1
      na_row$val <- NA
      new_rows[[length(new_rows) + 1]] <- na_row
    }
  }
  new_rows[[length(new_rows) + 1]] <- df[nrow(df), ]
  do.call(rbind, new_rows)
}

# Fallback/standard theme_pub function for consistent styling
theme_pub <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 1, color = "#1a1a1a"),
      plot.subtitle = element_text(size = base_size - 1, color = "#666666"),
      axis.title = element_text(size = base_size - 1, color = "#444444"),
      axis.text = element_text(size = base_size - 2, color = "#222222"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#ebebeb", size = 0.3)
    )
}
```

# 1. Executive Summary & Core Thesis

Fecal Indicator Bacteria (FIB) datasets are our primary tool for protecting public health from waterborne pathogens (see [WHO 2021 Guidelines for Recreational Water Quality](https://www.who.int/publications/i/item/9789240019041)). Focusing on **inland freshwaters** — rivers, lakes, and reservoirs, where *E. coli* is the primary indicator — we compiled a harmonized global dataset and find the monitoring infrastructure critically flawed in two distinct dimensions:

1. **Global Spatial Sparsity & Access Barriers:** Fecal pollution monitoring is highly unequal. Of the **5.31 million** clean freshwater records compiled, **~63%** are from the United States and **~26%** from Europe and Canada — leaving only **~11%** for the rest of the world (representing a significant gap in global water quality databases; see [UNEP 2021 SDG 6.3.2 Progress Report](https://www.unwater.org/publications/progress-water-quality-sdg-6-indicator-632-2021-updates)).
   * *The Sharing and Access Gap:* Crucially, this geographic disparity reflects not just where monitoring occurs, but **where data is shared in open, public, and machine-readable formats**. Large nations like China, Russia, and several countries in South America maintain extensive national and regional water quality monitoring networks. However, their high-resolution grab-sample data is frequently not published in, or harvested by, public, harmonized global repositories (e.g., [UNEP GEMStat](https://gemstat.org/)). Thus, we face a combination of a *collection gap* in developing or resource-limited nations and a *sharing/access barrier* in major industrial nations, rather than a total absence of monitoring.
2. **Temporal Mismatch ("Regulatory Rigidity"):** Even where data is abundant, it is rarely **actionable** for real-time safety. Water quality safety fluctuates rapidly, with bacterial concentrations spiking during rain events ("first flush") and decaying within 24–48 hours (see [Boehm et al., 2002 on short-term FIB variability](https://doi.org/10.1021/es011406c)). Yet, **only ~7% of sites (with ≥2 samples) achieve daily or near-daily sampling**. Instead, monitoring is structurally rigid, collected at weekly, monthly, or quarterly intervals, often restricted to summer bathing seasons. This regulatory-driven sampling design is blind to acute public health risks and fails to provide real-time safety metrics for drinking or recreation.

---

# Methods: Global FIB Data Compilation & Harmonization

> Outline draft for prose development. Full reference detail lives in
> `methods_digest.md`; the complete machine-readable source inventory (including
> excluded sources and reasons) is in `data_sources_summary.csv`.

### Overview
* This study is **freshwater-focused.** We assembled a harmonized global database of fecal
  indicator bacteria (FIB) concentrations from **nine public point-observation networks plus
  one annual-summary source**, standardized to a common schema and unit (**CFU/100 mL**).
  After cleaning and restriction to inland freshwaters, the analysis dataset comprises
  **5,314,791 individual observations** (`fecal_indicators_clean.feather`).
* ***E. coli* is the appropriate primary indicator for freshwater** recreational and drinking
  exposure (following [US EPA 2012 Recreational Water Quality Criteria](https://www.epa.gov/wqc/recreational-water-quality-criteria) and [EU Bathing Water Directive (2006/7/EC)](https://environment.ec.europa.eu/topics/water/bathing-water_en)), and it is the most
  abundant indicator here (~43% of freshwater records). Other co-reported indicators (fecal
  coliform, total coliform, intestinal enterococci, fecal streptococci) are retained where
  available, which is crucial as they collectively inform the broader FIB context. **Marine/coastal bathing waters — and the enterococci-based monitoring that
  dominates them — are out of scope by design** (see M.4).
* **By ecosystem**, the freshwater data are dominated by rivers/streams (72.4%) and lakes
  (22.6%), with reservoirs (2.1%), canals (1.9%), and ditches, springs, wetlands, and ponds
  making up the remainder.

### M.1 Data sources
Counts below are **freshwater observations** contributed by each source after cleaning
(full-database counts, including marine, are in `methods_digest.md`).

| Source | Region | Freshwater obs | Access |
|---|---|--:|---|
| WQP (US Water Quality Portal) | United States | 3,321,490 | API (`dataRetrieval`) |
| Eionet (EEA Bathing Water Directive) | EU/EEA | 763,440 | Web/CDR scrape + XML/GML |
| GEMStat (UN GEMS/Water) | Global (ex-US) | 399,706 | Manual (Zenodo release) |
| DataStream | Canada | 296,913 | API (`datastreamr`) |
| UK Open WIMS (EA Water Quality Archive) | United Kingdom | 189,485 | REST API |
| NMMP | South Africa | 161,467 | Manual CSV |
| Hub'Eau (Naïades) | France | 146,650 | API (`hubeau`) |
| LAWA | New Zealand | 35,248 | Manual Excel |
| WPdx (Water Point Data Exchange) | Global water points | 392 | Manual CSV |
| ANA *(annual summaries only)* | Brazil | 14,853 | Manual CSV → separate table |

* Five sources required **manual download** (GEMStat, NMMP, WPdx, LAWA, ANA); the rest
  were retrieved programmatically. Acquisition scripts are versioned in the companion
  data-pipeline repository.

### M.2 Indicator mix (freshwater)
* Within the freshwater dataset, no single indicator is dominant:
  **E. coli 2,308,643 (43.4%)**, fecal coliform 1,692,665 (31.8%), total coliform
  560,355 (10.5%), intestinal enterococci 479,053 (9.0%), fecal streptococci
  274,075 (5.2%).
* Retrieval was **prioritized around *E. coli***; other indicators were captured where
  co-reported, so per-source coverage of non–E. coli indicators is uneven (e.g., enterococci
  here come almost entirely from European inland bathing waters). Because the study is
  freshwater- and E. coli–anchored, this is a scope choice rather than a coverage gap — but
  cross-indicator comparisons of *total* monitoring effort should still be made cautiously, recognizing that these diverse parameters are all key components of Fecal Indicator Bacteria (FIB) monitoring frameworks.

### M.3 Harmonization
* **Units:** all concentrations converted to **CFU/100 mL**; original unit retained
  (`orig_unit`). **MPN is treated as CFU** (an approximation), **qPCR equivalents and
  non-convertible units (mass, turbidity, etc.) are dropped**.
* **Censored values:** detection-limit handling records direction in `censor_side`
  (left-censored → ½ × lower limit; right-censored → upper limit).
* **Location type:** source-specific descriptors mapped to a controlled vocabulary
  (river/stream, lake, ocean, estuary, canal, ditch, spring, pond, wetland, reservoir,
  groundwater/well, other).
* **Deduplication:** within-source averaging of repeated (site × date × time × indicator)
  records, then **cross-source fuzzy de-duplication** keyed on
  `date | indicator | value | ~1 km lat/lon`, retaining the most metadata-complete record.
* Default state backfill: WQP→"USA", DataStream→"Canada" where missing.

### M.4 Cleaning & freshwater restriction (`clean_data.R`)
* From the 9,055,019 harmonized records, cleaning keeps the canonical FIB keep-list and
  removes invalid dates (outside 1950–2026), out-of-range/`(0,0)` coordinates, and
  negative/extreme values (>10⁷ CFU/100 mL), leaving **9,044,108 cleaned records**.
* **Freshwater restriction.** Each record is classified as marine, groundwater, or
  freshwater. **Marine** is flagged by `loc_type` (ocean/estuary) **or** by the Eionet
  bathing-water category `BWCat` (`C` = coastal, `T` = transitional); `BWCat` overrides an
  ambiguous `loc_type`, which prevents coastal sites from hiding in the "other/unknown"
  bucket. **Groundwater/well** is excluded (distinct exposure pathway). This removes
  **3,409,529 marine** and **26,284 groundwater** records (→ 5,608,295).
* **Other/unknown resolution.** 506,058 records arrived with `loc_type` = other/NA; the
  marine guard leaves only `BWCat` L/R/NA here (no coastal leak). We **reclassify by
  `BWCat`** — `L`→lake (161,801) and `R`→river/stream (50,753), i.e. 212,554 European
  inland bathing waters relabeled — and **drop the 293,504 records that carry neither a
  usable `loc_type` nor a `BWCat`**, as they cannot be verified as inland. The final dataset
  is **5,314,791 freshwater records** (≈59% of the cleaned data), with no remaining
  'other'/unknown `loc_type`.

### M.5 Geographic distribution (freshwater)
* The freshwater data remain dominated by a few regions: **United States ≈ 62.9%**,
  **Europe ≈ 20.7%**, **Canada ≈ 5.7%**, **rest of world ≈ 10.7%** (≈89.3% from the
  US, Europe, and Canada combined).
* Restricting to freshwater **raises** the relative Global-South share (10.7% vs. ~6.9%
  across all water types), because Europe's contribution is heavily *coastal* bathing-water
  data that drops out. Leading non-US/Europe/Canada contributors: Mexico (256k), South
  Africa (161k), India (106k), New Zealand (35k).
* *(The Executive Summary now reports these freshwater shares. The earlier "54.3% US /
  43.7% Europe & Canada" all-water figures mis-assigned global GEMStat records to Europe —
  see `outline_review_flags.md`.)*

### M.6 Excluded sources, gaps, and the access spectrum
* A full accounting of attempted-but-excluded sources is in `data_sources_summary.csv`.
  Excluded for **known reasons** include Australia (Queensland; regional data access barriers), China (ESSD;
  provides monthly aggregates rather than raw daily/weekly grab-sample concentrations), India (CPCB; national portal access barriers + pre-aggregated tables), and EEA pre-2006
  bathing water (archived compliance classifications, not numeric concentrations). Colombia (IDEAM) and the APMP
  database were identified but **not pursued (reason undocumented)**.
* **Brazil (ANA) represents a key formatting/access barrier:** active monitoring occurs and sub-annual high-frequency sampling exists, but public data portals only publish aggregated annual statistics (minimum, maximum, and median summaries). While we compiled these in a separate table (`harmonized_annual.feather`), the raw, individual grab samples cannot be integrated into per-observation daily/weekly analyses, showing how data aggregation limits secondary research. WPdx similarly contributes sparse, fragmentary
  Global-South water-point data. Both cases illustrate that *open-access public availability*, not just
  *monitoring effort*, shapes the global picture.

---

# Figure 1: Global Distribution and Ecosystem Footprint

### Conceptual Description
* **Aims:** Establish the geographic inequality of global FIB monitoring, highlighting the data gaps (both collection and sharing gaps) and showcasing the types of aquatic ecosystems sampled.
* **Visual Components:**
  * **Main Map:** A global map using the **Robinson projection** (a clean, compromise projection that avoids the extreme polar distortion of Mercator). Observation counts are aggregated into a **hexagonal grid** density map to prevent single-station overlap and show regional density.
* **Key Findings:**
  * Rivers/streams dominate freshwater monitoring (**≈3.99 million records, ~75%**), followed by lakes/reservoirs (**≈1.33 million records, ~25%**).
  * The absence of records in open databases for large nations like China and Russia highlights the global data sharing and database integration barrier rather than a lack of monitoring activities in those nations.

### Executable R Code for Figure 1

```{r figure_1, fig.width=12, fig.height=5.2}
# Load the cleaned dataset
clean_data <- read_feather("fecal_indicators_clean.feather", as_data_frame = FALSE)

# Pull coordinates for mapping
coords_data <- clean_data %>%
  select(lat, lon) %>%
  filter(!is.na(lat) & !is.na(lon)) %>%
  collect()

# Get world map outline in Robinson projection (excluding Antarctica)
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(continent != "Antarctica")
world_robin <- st_transform(world, crs = "+proj=robin")

# Project coordinates to Robinson projection for hexbinning
coords <- as.matrix(coords_data[, c("lon", "lat")])
# Clean coords to ensure they are within bounds and exclude Antarctica (lat < -60)
valid_idx <- coords[, 1] >= -180 & coords[, 1] <= 180 & coords[, 2] >= -60 & coords[, 2] <= 90
coords_clean <- coords[valid_idx, ]
proj_coords <- sf_project(
  from = "+proj=longlat +datum=WGS84",
  to = "+proj=robin",
  pts = coords_clean
)
proj_df <- data.frame(x = proj_coords[, 1], y = proj_coords[, 2])

# Main map plot
p_map <- ggplot() +
  geom_sf(data = world_robin, fill = "#f5f5f7", color = "#ffffff", size = 0.2) +
  geom_hex(data = proj_df, aes(x = x, y = y), bins = 220, alpha = 0.85) +
  scale_fill_viridis_c(
    option = "mako", 
    trans = "log10", 
    direction = -1, 
    name = "Observations",
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  coord_sf(ylim = c(-5950000, 8625155), expand = FALSE) +
  theme_void() +
  labs(
    title = "Global Freshwater Fecal Indicator Bacteria (FIB) Data Density",
    subtitle = "Severe geographic disparities driven by both monitoring infrastructure and public data sharing"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5, color = "#2d2d2d"),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "#666666", margin = margin(b = 10)),
    legend.position = c(0.15, 0.28),
    legend.background = element_rect(fill = ggplot2::alpha("white", 0.7), color = NA),
    legend.key.width = unit(1.0, "cm")
  )

print(p_map)
```

---

# 2. Figure 2: Sampling Objectives & Mismatched Frequencies

### Conceptual Description
* **Aims:** Demonstrate that even within the same geographic region (Wisconsin), sampling designs are fragmented and serve different objectives. This shows the contrast between daily recreational monitoring, monthly regulatory compliance, and quarterly trend analysis.
* **Visual Components:**
  * **Left Panel (Map):** A zoomed-in map of Wisconsin plotting three freshwater sites monitored at contrasting frequencies:
    1. **Swimming/Recreation Site (Daily):** Wisconsin DNR inland lake beach (swimming-season daily sampling).
    2. **Regulatory Attainment Site (Monthly):** Wisconsin DNR river compliance site.
    3. **Long-Term Trend Site (Quarterly):** Wisconsin DNR river trend site.
  * **Right Panel (Timeseries):** Stacked timeseries plots showing the actual observation intervals for these three sites in 2022–2023.
* **Key Findings & Concept:**
  * **Swimming (Daily):** Protecting swimmers from pathogen exposure requires high-frequency (daily) monitoring to capture transient beach closures (see [Wade et al., 2003 on gastrointestinal illness and recreational water guidelines](https://doi.org/10.1289/ehp.6141)).
  * **Regulatory Compliance (Monthly):** Checking if a waterbody meets Clean Water Act water quality standards is typically satisfied by monthly regulatory grabs.
  * **Long-Term Trends (Quarterly):** Tracking multi-year effects of climate change or agricultural policy can be satisfied by low-frequency quarterly samples.
  * Tying these frequencies to different management objectives highlights why "abundant data" (which is mostly monthly/quarterly) cannot be used to protect recreational swimmers.

### Executable R Code for Figure 2

```{r figure_2, fig.width=12, fig.height=7}
# Load WI map boundaries
# Load map boundaries for USA states and filter Wisconsin to highlight it
states <- ne_states(country = "United States of America", returnclass = "sf")
wi_map <- states %>% filter(name == "Wisconsin")

# Download medium scale physical lakes and rivers
lakes_50m <- ne_download(scale = "medium", type = "lakes", category = "physical", returnclass = "sf")
rivers_50m <- ne_download(scale = "medium", type = "rivers_lake_centerlines", category = "physical", returnclass = "sf")

# Define major cities in Wisconsin for geographic reference
cities <- data.frame(
  name = c("Milwaukee", "Madison", "Green Bay"),
  lat = c(43.0389, 43.0731, 44.5133),
  lon = c(-87.9065, -89.4012, -88.0133)
)
cities_sf <- st_as_sf(cities, coords = c("lon", "lat"), crs = 4326)

# Define three representative freshwater sites in Wisconsin (Wisconsin DNR)
site_coords <- data.frame(
  site_id = c("WIDNR_WQX-683422", "WIDNR_WQX-543001", "WIDNR_WQX-313038"),
  label = c("1) Swimming (Daily Req.)", "2) Regulatory (Monthly Req.)", "3) Long-Term Trend (Quarterly Req.)"),
  lat = c(42.9374, 42.6091, 44.4585),
  lon = c(-88.4797, -89.0706, -87.5561)
)
site_coords_sf <- st_as_sf(site_coords, coords = c("lon", "lat"), crs = 4326)

# Plot zoomed-in Wisconsin map with surrounding states, hydrography, and major cities
p_ca_map <- ggplot() +
  geom_sf(data = states, fill = "#f5f5f7", color = "#e5e5ea", size = 0.2) +
  geom_sf(data = wi_map, fill = "#ffffff", color = "#4a4a4a", size = 0.5) +
  geom_sf(data = lakes_50m, fill = "#e8f2ff", color = "#a6c8e0", size = 0.15) +
  geom_sf(data = rivers_50m, color = "#a6c8e0", size = 0.3) +
  geom_sf(data = cities_sf, color = "#4a4a4a", size = 1.5) +
  geom_sf_text(data = cities_sf, aes(label = name), nudge_y = 0.12, size = 3, color = "#4a4a4a", fontface = "italic") +
  geom_sf(data = site_coords_sf, aes(color = label, shape = label), size = 5) +
  scale_color_manual(
    values = c(
      "1) Swimming (Daily Req.)" = "#de2d26", 
      "2) Regulatory (Monthly Req.)" = "#31a354", 
      "3) Long-Term Trend (Quarterly Req.)" = "#2b8cbe"
    ),
    name = "Monitoring Objective"
  ) +
  scale_shape_manual(
    values = c(
      "1) Swimming (Daily Req.)" = 16, 
      "2) Regulatory (Monthly Req.)" = 17, 
      "3) Long-Term Trend (Quarterly Req.)" = 15
    ),
    name = "Monitoring Objective"
  ) +
  coord_sf(xlim = c(-93.0, -86.5), ylim = c(42.3, 47.2), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.box.background = element_rect(color = "#d9d9d9", size = 0.5),
    panel.background = element_rect(fill = "#e8f2ff", color = NA), # Great Lakes / water fill
    plot.title = element_text(face = "bold", size = 12, color = "#2d2d2d")
  ) +
  labs(
    title = "A) Wisconsin Sites",
    x = "Longitude", y = "Latitude"
  )

# Extract observations for 2022-2023 for these three sites
ts_data <- clean_data %>%
  filter(site_id %in% c("WIDNR_WQX-683422", "WIDNR_WQX-543001", "WIDNR_WQX-313038")) %>%
  collect() %>%
  filter(date >= as.Date("2022-01-01") & date <= as.Date("2023-12-31")) %>%
  mutate(
    label = case_when(
      site_id == "WIDNR_WQX-683422" ~ "1) Swimming (Daily Req.)",
      site_id == "WIDNR_WQX-543001" ~ "2) Regulatory (Monthly Req.)",
      site_id == "WIDNR_WQX-313038" ~ "3) Long-Term Trend (Quarterly Req.)"
    )
  )

# Define helper function to insert NAs to break lines over winter/off-season gaps
insert_gaps <- function(df, max_gap) {
  if (nrow(df) < 2) return(df)
  df <- df[order(df$date), ]
  new_rows <- list()
  for (i in 1:(nrow(df) - 1)) {
    new_rows[[length(new_rows) + 1]] <- df[i, ]
    gap <- as.numeric(df$date[i+1] - df$date[i])
    if (gap > max_gap) {
      na_row <- df[i, ]
      na_row$date <- df$date[i] + 1
      na_row$val <- NA
      new_rows[[length(new_rows) + 1]] <- na_row
    }
  }
  new_rows[[length(new_rows) + 1]] <- df[nrow(df), ]
  do.call(rbind, new_rows)
}

# Apply insert_gaps with site-specific thresholds to avoid incorrect connecting lines
ts_data_with_na <- ts_data %>%
  group_by(label) %>%
  group_split() %>%
  lapply(function(df) {
    label_val <- df$label[1]
    max_gap <- case_when(
      label_val == "1) Swimming (Daily Req.)" ~ 15,
      label_val == "2) Regulatory (Monthly Req.)" ~ 45,
      label_val == "3) Long-Term Trend (Quarterly Req.)" ~ 120,
      TRUE ~ 30
    )
    insert_gaps(df, max_gap)
  }) %>%
  bind_rows()

# Generate stacked timeseries plot with broken lines for gaps
p_ts <- ggplot(ts_data_with_na, aes(x = date, y = val, color = label, shape = label)) +
  geom_line(size = 0.4, alpha = 0.5, na.rm = TRUE) +
  geom_point(size = 2, na.rm = TRUE) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(
    values = c(
      "1) Swimming (Daily Req.)" = "#de2d26", 
      "2) Regulatory (Monthly Req.)" = "#31a354", 
      "3) Long-Term Trend (Quarterly Req.)" = "#2b8cbe"
    )
  ) +
  scale_shape_manual(
    values = c(
      "1) Swimming (Daily Req.)" = 16, 
      "2) Regulatory (Monthly Req.)" = 17, 
      "3) Long-Term Trend (Quarterly Req.)" = 15
    )
  ) +
  facet_wrap(~label, ncol = 1, scales = "free_y") +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 9, hjust = 0),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12, color = "#2d2d2d")
  ) +
  labs(
    title = "B) Observations (2022-2023)",
    x = "Date",
    y = "E. coli Concentration (CFU/100mL)"
  )

# Combine map and timeseries side-by-side
p_fig2_combined <- p_ca_map + p_ts + plot_layout(widths = c(1, 1.3))
print(p_fig2_combined)
```

---

# 3. Figure 3: Quantifying the Actionability Gap

### Conceptual Description
* **Aims:** Provide a quantitative summary of the data density bottleneck and sampling frequency limitations across the entire global dataset.
* **Visual Components:**
  * **Panel A (Data Volume):** A cumulative distribution function (CDF) or step-survival curve showing the percentage of sites that survive as the minimum observation threshold ($N$) increases from 1 to 1,000.
  * **Panel B (Sampling Interval):** A bar plot of the average sampling return interval (Daily, Weekly, Monthly, Quarterly, Yearly, $> \text{Yearly}$) for all multi-sample sites.
* **Key Findings:**
  * Most sites have very few observations: **only 10.8% of sites have $\ge 100$ observations**, and only **0.7% have $\ge 500$ observations** (see [WHO Guidelines for Recreational Water Quality](https://www.who.int/publications/i/item/9789240019041)).
  * Among sites with at least 2 samples, **only 7.0% are monitored on a daily or near-daily basis** ($\le 1.5$ days). The vast majority are monitored **monthly (29.9%)** or **quarterly (30.7%)**, which fails to match the fast-changing kinetics of waterborne pathogen spikes (see [Boehm et al., 2002](https://doi.org/10.1021/es011406c)).

### Executable R Code for Figure 3

```{r figure_3, fig.width=12, fig.height=6}
# Group by site and calculate observation counts and spans
site_summary <- clean_data %>%
  select(site_id, lat, lon, date) %>%
  collect() %>%
  mutate(
    site_key = if_else(is.na(site_id), paste0("coord_", lat, "_", lon), site_id)
  ) %>%
  group_by(site_key) %>%
  summarise(
    n_obs = n(),
    min_date = min(date),
    max_date = max(date),
    span_days = as.numeric(max_date - min_date),
    .groups = "drop"
  )

# Part A: Threshold survival curve
thresholds <- c(1, 2, 5, 10, 20, 50, 100, 250, 500, 1000)
survival_df <- data.frame(
  threshold = thresholds,
  sites = sapply(thresholds, function(t) sum(site_summary$n_obs >= t))
) %>%
  mutate(pct = (sites / nrow(site_summary)) * 100)

p_survival <- ggplot(survival_df, aes(x = factor(threshold), y = pct, group = 1)) +
  geom_area(fill = "#3182bd", alpha = 0.12) +
  geom_line(color = "#3182bd", linewidth = 1.1) +
  geom_point(color = "#08519c", size = 2.6) +
  geom_text(aes(label = paste0(round(pct), "%")), vjust = -1, size = 2.8, color = "#08519c") +
  expand_limits(y = 108) +
  theme_pub(11) +
  labs(
    title = "A) Site Data Volume Survival Curve",
    x = "Minimum Observations per Site",
    y = "Percentage of Sites (%)"
  )

# Part B: Sampling interval distribution
interval_df <- site_summary %>%
  filter(n_obs >= 2) %>%
  mutate(
    avg_interval = span_days / (n_obs - 1)
  ) %>%
  mutate(
    interval_cat = case_when(
      avg_interval <= 1.5 ~ "Daily\n(<=1.5d)",
      avg_interval > 1.5 & avg_interval <= 8 ~ "Weekly\n(1.5-8d)",
      avg_interval > 8 & avg_interval <= 31 ~ "Monthly\n(8-31d)",
      avg_interval > 31 & avg_interval <= 92 ~ "Quarterly\n(31-92d)",
      avg_interval > 92 & avg_interval <= 366 ~ "Yearly\n(92-366d)",
      avg_interval > 366 ~ "Longer\n(>366d)",
      TRUE ~ "Unknown"
    )
  ) %>%
  count(interval_cat) %>%
  mutate(pct = (n / sum(n)) * 100)

# Reorder categories logically
interval_df$interval_cat <- factor(
  interval_df$interval_cat, 
  levels = c("Daily\n(<=1.5d)", "Weekly\n(1.5-8d)", "Monthly\n(8-31d)", "Quarterly\n(31-92d)", "Yearly\n(92-366d)", "Longer\n(>366d)")
)

# Highlight the policy-relevant "Daily" category (the key finding: very few sites)
interval_df$highlight <- ifelse(grepl("Daily", interval_df$interval_cat), "Daily", "Other")

p_intervals <- ggplot(interval_df, aes(x = interval_cat, y = pct, fill = highlight)) +
  geom_col(show.legend = FALSE, width = 0.72) +
  geom_text(aes(label = paste0(round(pct, 1), "%")), vjust = -0.5, size = 3, color = "#2d2d2d") +
  scale_fill_manual(values = c("Daily" = "#d7301f", "Other" = "#74a9cf")) +
  expand_limits(y = max(interval_df$pct) * 1.13) +
  theme_pub(11) +
  theme(panel.grid.major.x = element_blank()) +
  labs(
    title = "B) Distribution of Sampling Return Intervals",
    x = "Sampling Interval Category",
    y = "Percentage of Multi-Sample Sites (%)"
  )

# Combine plots side-by-side
p_fig3_combined <- p_survival + p_intervals
print(p_fig3_combined)
```

---

# Figure 4: Reclaiming Recreation Days through High-Frequency Monitoring

### Conceptual Description
* **Aims:** Demonstrate that bacterial pollution in recreational waters is transient and that high-frequency data from across the globe allows precise management.
* **Visual Components:**
  * **Panel A (Map):** A global map in Robinson projection showing the location of three high-frequency case studies: Euclid Beach (USA), River Lune (UK), and Taruheru River (New Zealand).
  * **Panel B (Timeseries):** A 3-column timeseries layout showing the high-frequency variation of E. coli in 2023 for all three sites relative to the safety limit of 235 CFU/100mL, highlighting short-lived spikes and rapid recovery.
* **Key Findings & Concept:**
  * Pollution is a passing state, not a permanent attribute of most beaches or rivers.
  * With monthly sampling, a single bad sample can impair a beach for an entire month. High-frequency daily sampling allows managers to target closures precisely (only closing during short-lived spikes) and keep the waters open for recreation on the remaining safe days (see [populationmedicine.eu's paper on daily vs weekly sampling and predictive models](https://doi.org/10.18332/popmed/163820)).

### Executable R Code for Figure 4

```{r figure_4, fig.width=12, fig.height=8.5}
# Project case study sites for mapping
case_sites <- data.frame(
  site_id = c("21OHBCH-OHEUSP01", "uk_bwq_NW-88003843", "lawa_gdc-00033"),
  label = c("A", "B", "C"),
  lat = c(41.5843, 53.8590, -38.66849),
  lon = c(-81.5686, -2.7436, 178.0286)
)

# Project to Robinson coordinates
coords_case <- as.matrix(case_sites[, c("lon", "lat")])
proj_coords_case <- sf_project(
  from = "+proj=longlat +datum=WGS84",
  to = "+proj=robin",
  pts = coords_case
)
case_sites$x <- proj_coords_case[, 1]
case_sites$y <- proj_coords_case[, 2]

# Add specific nudges to labels to prevent overlap with markers or borders
case_sites <- case_sites %>%
  mutate(
    nudge_x = case_when(
      site_id == "21OHBCH-OHEUSP01" ~ -800000,
      site_id == "uk_bwq_NW-88003843" ~ 800000,
      site_id == "lawa_gdc-00033" ~ 1200000
    ),
    nudge_y = case_when(
      site_id == "21OHBCH-OHEUSP01" ~ 500000,
      site_id == "uk_bwq_NW-88003843" ~ 500000,
      site_id == "lawa_gdc-00033" ~ 0
    )
  )

# Get world map outline in Robinson projection (excluding Antarctica)
world_robin_fig4 <- st_transform(
  ne_countries(scale = "medium", returnclass = "sf") %>% filter(continent != "Antarctica"),
  crs = "+proj=robin"
)

# Global map plot showing site locations (trimmed bottom coordinates)
p_map_case <- ggplot() +
  geom_sf(data = world_robin_fig4, fill = "#f5f5f7", color = "#ffffff", size = 0.2) +
  geom_point(data = case_sites, aes(x = x, y = y), color = "#d7301f", size = 4, shape = 18) +
  geom_text(data = case_sites, aes(x = x + nudge_x, y = y + nudge_y, label = label), 
            fontface = "bold", size = 5, color = "#2d2d2d") +
  coord_sf(ylim = c(-5950000, 8625155), expand = FALSE) +
  theme_void() +
  labs(
    title = "A) Global High-Frequency Case Studies",
    subtitle = "Locations of selected E. coli monitoring sites analyzed below"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5, color = "#2d2d2d"),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "#666666", margin = margin(b = 5))
  )

# Get timeseries data for the three sites in 2023
ts_case_data <- clean_data %>%
  filter(site_id %in% c("21OHBCH-OHEUSP01", "uk_bwq_NW-88003843", "lawa_gdc-00033")) %>%
  collect() %>%
  filter(lubridate::year(date) == 2023)

# US Beach Timeseries (polished styling)
ts_us <- ts_case_data %>% 
  filter(site_id == "21OHBCH-OHEUSP01") %>%
  insert_gaps(30)

p_ts_us <- ggplot(ts_us, aes(x = date, y = val)) +
  annotate("rect", xmin = min(ts_us$date, na.rm = TRUE), xmax = max(ts_us$date, na.rm = TRUE),
           ymin = 1, ymax = 235, fill = "#2ca25f", alpha = 0.08) +
  geom_hline(yintercept = 235, linetype = "dashed", color = "#d7301f", linewidth = 0.7) +
  geom_line(color = "grey75", linewidth = 0.4, na.rm = TRUE) +
  geom_point(aes(color = val > 235), size = 2.3, na.rm = TRUE) +
  scale_color_manual(values = c("FALSE" = "#2ca25f", "TRUE" = "#d7301f")) +
  scale_y_log10(labels = scales::comma) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  theme_pub(11) +
  theme(legend.position = "none") +
  labs(
    title = "B) Case A: Euclid Beach, USA",
    x = NULL,
    y = "E. coli (CFU/100mL) [Log Scale]"
  )

# UK River Timeseries (polished styling)
ts_uk <- ts_case_data %>% 
  filter(site_id == "uk_bwq_NW-88003843") %>%
  insert_gaps(30)

p_ts_uk <- ggplot(ts_uk, aes(x = date, y = val)) +
  annotate("rect", xmin = min(ts_uk$date, na.rm = TRUE), xmax = max(ts_uk$date, na.rm = TRUE),
           ymin = 1, ymax = 235, fill = "#2ca25f", alpha = 0.08) +
  geom_hline(yintercept = 235, linetype = "dashed", color = "#d7301f", linewidth = 0.7) +
  geom_line(color = "grey75", linewidth = 0.4, na.rm = TRUE) +
  geom_point(aes(color = val > 235), size = 2.3, na.rm = TRUE) +
  scale_color_manual(values = c("FALSE" = "#2ca25f", "TRUE" = "#d7301f")) +
  scale_y_log10(labels = scales::comma) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  theme_pub(11) +
  theme(legend.position = "none") +
  labs(
    title = "C) Case B: River Lune, UK",
    x = NULL,
    y = NULL
  )

# NZ River Timeseries (polished styling)
ts_nz <- ts_case_data %>% 
  filter(site_id == "lawa_gdc-00033") %>%
  insert_gaps(30)

p_ts_nz <- ggplot(ts_nz, aes(x = date, y = val)) +
  annotate("rect", xmin = min(ts_nz$date, na.rm = TRUE), xmax = max(ts_nz$date, na.rm = TRUE),
           ymin = 1, ymax = 235, fill = "#2ca25f", alpha = 0.08) +
  geom_hline(yintercept = 235, linetype = "dashed", color = "#d7301f", linewidth = 0.7) +
  geom_line(color = "grey75", linewidth = 0.4, na.rm = TRUE) +
  geom_point(aes(color = val > 235), size = 2.3, na.rm = TRUE) +
  scale_color_manual(values = c("FALSE" = "#2ca25f", "TRUE" = "#d7301f")) +
  scale_y_log10(labels = scales::comma) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  theme_pub(11) +
  theme(legend.position = "none") +
  labs(
    title = "D) Case C: Taruheru River, NZ",
    x = NULL,
    y = NULL
  )

# Combine plots: Map at top, 3 timeseries side-by-side at bottom
p_fig4_combined <- p_map_case / (p_ts_us | p_ts_uk | p_ts_nz) + plot_layout(heights = c(1.2, 1.0))
print(p_fig4_combined)
```
```

---

# Figure 5: Global Safety of High-Frequency Monitoring Sites

### Conceptual Description
* **Aims:** Map the global distribution of high-frequency E. coli monitoring sites (with at least 2 years of daily/near-daily data) and visualize their overall safety records.
* **Visual Components:**
  * **Panel A (Global Map):** A global Robinson projection map plotting high-frequency E. coli sites, with points colored by the individual exceedance rate (percentage of samples > 235 CFU/100mL) from green (0%) to red (100%).
  * **Panel B (Safety Donut Chart):** A donut chart summarizing the overall compliance rate across all high-frequency observations (77.6% safe vs 22.4% unsafe).
* **Key Findings:**
  * High-frequency monitoring is almost exclusively restricted to North America and Europe, with a few sites in South Africa and New Zealand (representing the open data access footprint rather than an absence of monitoring elsewhere).
  * Even among these sites, water quality varies significantly. Many sites are consistently safe (0-5% exceedance), but a substantial portion in agricultural or urban zones show chronic pollution (>20% exceedance), illustrating the spatial heterogeneity of fecal pollution.
  * Across all high-frequency daily monitoring observations (representing 631 high-density sites and 235k samples), the water is safe for contact recreation 77.6% of the time (see [US EPA 2012 Recreational Water Quality Criteria](https://www.epa.gov/wqc/recreational-water-quality-criteria) recommendations).

### Executable R Code for Figure 5

```{r figure_5, fig.width=12, fig.height=5.5}
# Filter and compute site summaries in Arrow and R
site_stats <- clean_data %>%
  filter(var == "Escherichia coli") %>%
  group_by(site_id, country, lat, lon) %>%
  summarise(
    n_obs = n(),
    min_date = min(date),
    max_date = max(date),
    span_days = as.numeric(max_date - min_date),
    .groups = "drop"
  ) %>%
  filter(n_obs >= 50 & span_days >= 730)

raw_dates <- clean_data %>%
  semi_join(site_stats, by = "site_id") %>%
  select(site_id, date, val) %>%
  collect()

median_gaps_exceed <- raw_dates %>%
  group_by(site_id) %>%
  arrange(date) %>%
  summarise(
    median_gap = median(as.numeric(diff(date), units = "days")),
    exceedance_pct = sum(val > 235) / n() * 100,
    .groups = "drop"
  )

final_sites <- site_stats %>%
  collect() %>%
  left_join(median_gaps_exceed, by = "site_id") %>%
  filter(median_gap <= 3)

# Project to Robinson projection
coords_fig5 <- as.matrix(final_sites[, c("lon", "lat")])
valid_idx_fig5 <- coords_fig5[, 1] >= -180 & coords_fig5[, 1] <= 180 & coords_fig5[, 2] >= -60 & coords_fig5[, 2] <= 90
coords_clean_fig5 <- coords_fig5[valid_idx_fig5, ]
proj_coords_fig5 <- sf_project(
  from = "+proj=longlat +datum=WGS84",
  to = "+proj=robin",
  pts = coords_clean_fig5
)
proj_df_fig5 <- data.frame(
  x = proj_coords_fig5[, 1], 
  y = proj_coords_fig5[, 2],
  exceedance_pct = final_sites$exceedance_pct[valid_idx_fig5]
)

# Load world map outline (excluding Antarctica)
world_robin_fig5 <- st_transform(
  ne_countries(scale = "medium", returnclass = "sf") %>% filter(continent != "Antarctica"),
  crs = "+proj=robin"
)

# Plot Figure 5 Map
p_map_fig5 <- ggplot() +
  geom_sf(data = world_robin_fig5, fill = "#f5f5f7", color = "#ffffff", size = 0.2) +
  geom_point(data = proj_df_fig5, aes(x = x, y = y, color = exceedance_pct), size = 0.8, alpha = 0.7) +
  scale_color_gradientn(
    colors = c("#2ca25f", "#e6f598", "#fee08b", "#fdae61", "#f46d43", "#d7301f"),
    name = "Exceedance Rate\n(% of samples > 235 CFU)"
  ) +
  coord_sf(ylim = c(-5950000, 8625155), expand = FALSE) +
  theme_void() +
  labs(
    title = "A) Global High-Frequency E. coli Sites",
    subtitle = "Exceedance rates at sites with >= 2 years of daily/near-daily data"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5, color = "#2d2d2d"),
    plot.subtitle = element_text(size = 9, hjust = 0.5, color = "#666666", margin = margin(b = 10)),
    legend.position = "bottom",
    legend.key.width = unit(1.0, "cm"),
    legend.title = element_text(size = 8, color = "#2d2d2d"),
    legend.text = element_text(size = 7, color = "#2d2d2d")
  )

# Create the donut chart
safe_df <- data.frame(
  Status = c("Safe (<235 CFU)", "Unsafe (>235 CFU)"),
  Percentage = c(77.6, 22.4)
)

p_safe <- ggplot(safe_df, aes(x = 2, y = Percentage, fill = Status)) +
  geom_col(width = 1, color = "white", linewidth = 1) +
  coord_polar(theta = "y", start = 0) +
  xlim(c(0.4, 2.5)) +
  geom_text(aes(label = paste0(Percentage, "%")),
            position = position_stack(vjust = 0.5), color = "white", fontface = "bold", size = 4.5) +
  annotate("text", x = 0.4, y = 0, label = "631 sites\n235k samples",
           size = 3.2, color = "#444444", lineheight = 0.95, fontface = "bold") +
  scale_fill_manual(values = c("Safe (<235 CFU)" = "#2ca25f", "Unsafe (>235 CFU)" = "#d7301f"),
                    name = NULL) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8, color = "#2d2d2d"),
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5, color = "#2d2d2d"),
    plot.subtitle = element_text(size = 9, hjust = 0.5, color = "#666666", margin = margin(b = 10))
  ) +
  labs(
    title = "B) Global Safety Summary",
    subtitle = "Recreational threshold compliance across all daily observations"
  )

# Combine map and donut chart side-by-side
p_fig5_combined <- p_map_fig5 + p_safe + plot_layout(widths = c(2.2, 1))
print(p_fig5_combined)
```
```

---

# 5. Discussion & Synthesis Plan for the Paper

### 1. The Public Data Access and Sharing Gap
* The geographic data gaps in open-access repositories for regions like China, Russia, and parts of South America are primarily driven by **portal formatting and database access barriers** rather than an absence of water quality monitoring (see [UNEP 2021 Progress on Water Quality](https://www.unwater.org/publications/progress-water-quality-sdg-6-indicator-632-2021-updates)). These nations maintain substantial agricultural, industrial, and urban water management networks that collect regular fecal indicator bacteria (FIB) data, but these resources are typically hosted on internal agency servers or offline portals rather than being syndicated to global databases.
* *Scientific Recommendation:* Resolving global water security requires international data synchronization standards similar to the World Meteorological Organization's unified policies (see [WMO Unified Data Policy Resolution 1](https://wmo.int/about-us/governance/data-policy)), enabling programmatic harvesting of water quality metrics.

### 2. The Public Health Mismatch (Swimming vs. Standards)
* Regulatory thresholds (such as the EPA limit of 235 CFU/100mL for E. coli) are designed to manage long-term exposure risks (see [US EPA 2012 Recreational Water Quality Criteria](https://www.epa.gov/wqc/recreational-water-quality-criteria)). However, applying these limits to low-frequency weekly or monthly samples leads to a **high rate of false negatives** (missing acute pathogen spikes after rainfall) and **false positives** (maintaining beach closures or impairments long after contaminants have cleared).
* Because traditional culture-based testing requires 24 hours of incubation, public health advisories are inherently **reactive**—representing yesterday's water safety rather than today's risk.
* **Delayed Attainment Updating under the Clean Water Act:** Under CWA Section 303(d), impairment and attainment statuses are formally updated and reported to the EPA only **biennially (every two years)**. Because this administrative cycle is slow, waterbodies can carry an "impaired" designation continuously. Yet, our high-frequency data demonstrates that many of these sites experience extended periods where FIB levels are low and safe for contact. Without dynamic testing and real-time public communication, static administrative listings can unnecessarily discourage recreation during safe intervals, while failing to protect users from sudden storm-driven pollution.

### 3. Policy and Future Research Recommendations
* **Event-Based Sampling:** Transition from calendar-rigid (weekly/monthly) sampling schedules to weather-driven, precipitation-triggered testing.
* **Predictive Modeling:** Integrate machine learning and regression-based modeling (using rain gauges, turbidity, and river discharge) to project recreational safety in real-time (similar to the [USGS Nowcast program](https://www.usgs.gov/mission-areas/water-resources/science/nowcast-predictive-models-beach-water-quality)).
* **Rapid Diagnostics:** Promote the adoption of qPCR molecular methods and automated enzymatic sensors to reduce the 24-hour incubation lag to under 3 hours (see [Griffith et al., 2012 on rapid molecular methods for beach management](https://doi.org/10.1016/j.watres.2011.11.042)).
