---
title: "Paper Outline: The Global Sparsity and Regulatory Rigidity of Fecal Indicator Bacteria Monitoring"
author: "Matthew R. V. Ross"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(arrow)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(sf)
library(rnaturalearth)
library(patchwork)
```

# 1. Executive Summary & Core Thesis

Fecal Indicator Bacteria (FIB) datasets are our primary tool for protecting public health from waterborne pathogens. Focusing on **inland freshwaters** — rivers, lakes, and reservoirs, where *E. coli* is the standard indicator — we compiled a harmonized global dataset and find the monitoring infrastructure critically flawed in two distinct dimensions:

1. **Global Spatial Sparsity & Access Barriers:** Fecal pollution monitoring is highly unequal. Of the **5.31 million** clean freshwater records compiled, **~63%** are from the United States and **~26%** from Europe and Canada — leaving only **~11%** for the rest of the world.
   * *The Sharing Gap:* Crucially, this geographic disparity reflects not just where monitoring occurs, but **where data is shared**. Large nations like Russia, China, and parts of South America almost certainly maintain national water quality networks, but their data is absent from public, harmonized global repositories. Thus, we face both a *collection gap* in developing nations and a *sharing gap* in major industrial nations.
2. **Temporal Mismatch ("Regulatory Rigidity"):** Even where data is abundant, it is rarely **actionable** for real-time safety. Water quality safety fluctuates rapidly, with bacterial concentrations spiking during rain events ("first flush") and decaying within 24–48 hours. Yet, **only ~7% of sites (with ≥2 samples) achieve daily or near-daily sampling**. Instead, monitoring is structurally rigid, collected at weekly, monthly, or quarterly intervals, often restricted to summer bathing seasons. This regulatory-driven sampling design is blind to acute public health risks and fails to provide real-time safety metrics for drinking or recreation.

---

# Methods: Global FIB Data Compilation & Harmonization

> Outline draft for prose development. Full reference detail lives in
> `methods_digest.md`; the complete machine-readable source inventory (including
> excluded sources and reasons) is in `data_sources_summary.csv`.

### Overview
* This study is **freshwater-focused.** We assembled a harmonized global database of fecal
  indicator bacteria (FIB) concentrations from **nine public point-observation networks plus
  one annual-summary source**, standardized to a common schema and unit (**CFU/100 mL**).
  After cleaning and restriction to inland freshwaters, the analysis dataset comprises
  **5,314,791 individual observations** (`fecal_indicators_clean.feather`).
* ***E. coli* is the appropriate primary indicator for freshwater** recreational and drinking
  exposure (US EPA freshwater criteria; EU inland bathing standards), and it is the most
  abundant indicator here (~43% of freshwater records). Other co-reported indicators (fecal
  coliform, total coliform, intestinal enterococci, fecal streptococci) are retained where
  available. **Marine/coastal bathing waters — and the enterococci-based monitoring that
  dominates them — are out of scope by design** (see M.4).
* **By ecosystem**, the freshwater data are dominated by rivers/streams (72.4%) and lakes
  (22.6%), with reservoirs (2.1%), canals (1.9%), and ditches, springs, wetlands, and ponds
  making up the remainder.

### M.1 Data sources
Counts below are **freshwater observations** contributed by each source after cleaning
(full-database counts, including marine, are in `methods_digest.md`).

| Source | Region | Freshwater obs | Access |
|---|---|--:|---|
| WQP (US Water Quality Portal) | United States | 3,321,490 | API (`dataRetrieval`) |
| Eionet (EEA Bathing Water Directive) | EU/EEA | 763,440 | Web/CDR scrape + XML/GML |
| GEMStat (UN GEMS/Water) | Global (ex-US) | 399,706 | Manual (Zenodo release) |
| DataStream | Canada | 296,913 | API (`datastreamr`) |
| UK Open WIMS (EA Water Quality Archive) | United Kingdom | 189,485 | REST API |
| NMMP | South Africa | 161,467 | Manual CSV |
| Hub'Eau (Naïades) | France | 146,650 | API (`hubeau`) |
| LAWA | New Zealand | 35,248 | Manual Excel |
| WPdx (Water Point Data Exchange) | Global water points | 392 | Manual CSV |
| ANA *(annual summaries only)* | Brazil | 14,853 | Manual CSV → separate table |

* Five sources required **manual download** (GEMStat, NMMP, WPdx, LAWA, ANA); the rest
  were retrieved programmatically. Acquisition scripts are versioned in the companion
  data-pipeline repository.

### M.2 Indicator mix (freshwater)
* Within the freshwater dataset, no single indicator is dominant:
  **E. coli 2,308,643 (43.4%)**, fecal coliform 1,692,665 (31.8%), total coliform
  560,355 (10.5%), intestinal enterococci 479,053 (9.0%), fecal streptococci
  274,075 (5.2%).
* Retrieval was **prioritized around *E. coli***; other indicators were captured where
  co-reported, so per-source coverage of non–E. coli indicators is uneven (e.g., enterococci
  here come almost entirely from European inland bathing waters). Because the study is
  freshwater- and E. coli–anchored, this is a scope choice rather than a coverage gap — but
  cross-indicator comparisons of *total* monitoring effort should still be made cautiously.

### M.3 Harmonization
* **Units:** all concentrations converted to **CFU/100 mL**; original unit retained
  (`orig_unit`). **MPN is treated as CFU** (an approximation), **qPCR equivalents and
  non-convertible units (mass, turbidity, etc.) are dropped**.
* **Censored values:** detection-limit handling records direction in `censor_side`
  (left-censored → ½ × lower limit; right-censored → upper limit).
* **Location type:** source-specific descriptors mapped to a controlled vocabulary
  (river/stream, lake, ocean, estuary, canal, ditch, spring, pond, wetland, reservoir,
  groundwater/well, other).
* **Deduplication:** within-source averaging of repeated (site × date × time × indicator)
  records, then **cross-source fuzzy de-duplication** keyed on
  `date | indicator | value | ~1 km lat/lon`, retaining the most metadata-complete record.
* Default state backfill: WQP→"USA", DataStream→"Canada" where missing.

### M.4 Cleaning & freshwater restriction (`clean_data.R`)
* From the 9,055,019 harmonized records, cleaning keeps the canonical FIB keep-list and
  removes invalid dates (outside 1950–2026), out-of-range/`(0,0)` coordinates, and
  negative/extreme values (>10⁷ CFU/100 mL), leaving **9,044,108 cleaned records**.
* **Freshwater restriction.** Each record is classified as marine, groundwater, or
  freshwater. **Marine** is flagged by `loc_type` (ocean/estuary) **or** by the Eionet
  bathing-water category `BWCat` (`C` = coastal, `T` = transitional); `BWCat` overrides an
  ambiguous `loc_type`, which prevents coastal sites from hiding in the "other/unknown"
  bucket. **Groundwater/well** is excluded (distinct exposure pathway). This removes
  **3,409,529 marine** and **26,284 groundwater** records (→ 5,608,295).
* **Other/unknown resolution.** 506,058 records arrived with `loc_type` = other/NA; the
  marine guard leaves only `BWCat` L/R/NA here (no coastal leak). We **reclassify by
  `BWCat`** — `L`→lake (161,801) and `R`→river/stream (50,753), i.e. 212,554 European
  inland bathing waters relabeled — and **drop the 293,504 records that carry neither a
  usable `loc_type` nor a `BWCat`**, as they cannot be verified as inland. The final dataset
  is **5,314,791 freshwater records** (≈59% of the cleaned data), with no remaining
  'other'/unknown `loc_type`.

### M.5 Geographic distribution (freshwater)
* The freshwater data remain dominated by a few regions: **United States ≈ 62.9%**,
  **Europe ≈ 20.7%**, **Canada ≈ 5.7%**, **rest of world ≈ 10.7%** (≈89.3% from the
  US, Europe, and Canada combined).
* Restricting to freshwater **raises** the relative Global-South share (10.7% vs. ~6.9%
  across all water types), because Europe's contribution is heavily *coastal* bathing-water
  data that drops out. Leading non-US/Europe/Canada contributors: Mexico (256k), South
  Africa (161k), India (106k), New Zealand (35k).
* *(The Executive Summary now reports these freshwater shares. The earlier "54.3% US /
  43.7% Europe & Canada" all-water figures mis-assigned global GEMStat records to Europe —
  see `outline_review_flags.md`.)*

### M.6 Excluded sources, gaps, and the access spectrum
* A full accounting of attempted-but-excluded sources is in `data_sources_summary.csv`.
  Excluded for **known reasons** include Australia (Queensland; access wall), China (ESSD;
  monthly aggregates only), India (CPCB; access wall + aggregation), and EEA pre-2006
  bathing water (compliance classes, not concentrations). Colombia (IDEAM) and the APMP
  database were identified but **not pursued (reason undocumented)**.
* **Brazil (ANA) is an instructive case of the *sharing gap*:** sub-annual sampling
  demonstrably exists, but the public-facing release provides **only annual min/max/median
  summaries**, which we retain in a separate `harmonized_annual.feather` table (they cannot
  enter the per-observation analyses). WPdx similarly contributes sparse, fragmentary
  Global-South water-point data. Both illustrate that *public availability*, not just
  *monitoring effort*, shapes the global picture.

---

# Figure 1: Global Distribution and Ecosystem Footprint

### Conceptual Description
* **Aims:** Establish the geographic inequality of global FIB monitoring, highlighting the data gaps (both collection and sharing gaps) and showcasing the types of aquatic ecosystems sampled.
* **Visual Components:**
  * **Main Map:** A global map using the **Robinson projection** (a clean, compromise projection that avoids the extreme polar distortion of Mercator). Observation counts are aggregated into a **hexagonal grid** density map to prevent single-station overlap and show regional density.
  * **Inset Sub-Plot:** A horizontal bar plot showing the total number of freshwater observations collapsed into two main ecosystem classes: **River/Stream** (incl. canals, ditches, springs) and **Lake/Reservoir** (incl. ponds, wetlands). This inset is placed in the northern/eastern region of Eurasia (Siberia/China) where there is a notable absence of publicly shared monitoring data.
* **Key Findings:**
  * Rivers/streams dominate freshwater monitoring (**≈3.99 million records, ~75%**), followed by lakes/reservoirs (**≈1.33 million records, ~25%**).
  * The visual void across Russia and China highlights the global data sharing barrier.

### Executable R Code for Figure 1

```{r figure_1, fig.width=12, fig.height=8}
# Load the cleaned dataset
clean_data <- read_feather("fecal_indicators_clean.feather", as_data_frame = FALSE)

# Collapse ecosystem types
ecosystem_data <- clean_data %>%
  select(loc_type, BWCat, lat, lon) %>%
  filter(!is.na(lat) & !is.na(lon)) %>%
  collect() %>%
  mutate(
    # Freshwater-only dataset: collapse loc_type into two ecosystem classes
    ecosystem = case_when(
      loc_type %in% c("river/stream", "canal", "ditch", "spring") ~ "River/Stream",
      loc_type %in% c("lake", "reservoir", "pond", "wetland") ~ "Lake/Reservoir",
      TRUE ~ "Other"
    )
  )

# Get world map outline in Robinson projection
world <- ne_countries(scale = "medium", returnclass = "sf")
world_robin <- st_transform(world, crs = "+proj=robin")

# Project coordinates to Robinson projection for hexbinning
coords <- as.matrix(ecosystem_data[, c("lon", "lat")])
# Clean coords to ensure they are within bounds
valid_idx <- coords[, 1] >= -180 & coords[, 1] <= 180 & coords[, 2] >= -90 & coords[, 2] <= 90
coords_clean <- coords[valid_idx, ]
proj_coords <- sf_project(
  from = "+proj=longlat +datum=WGS84",
  to = "+proj=robin",
  pts = coords_clean
)
proj_df <- data.frame(x = proj_coords[, 1], y = proj_coords[, 2])

# Main map plot
p_map <- ggplot() +
  geom_sf(data = world_robin, fill = "#f0f0f0", color = "#d9d9d9", size = 0.15) +
  geom_hex(data = proj_df, aes(x = x, y = y), bins = 90, alpha = 0.85) +
  scale_fill_viridis_c(
    option = "magma", 
    trans = "log10", 
    direction = -1, 
    name = "Observations\n(Log10 Count)"
  ) +
  theme_void() +
  labs(
    title = "Global Freshwater Fecal Indicator Bacteria (FIB) Data Density",
    subtitle = "Severe geographic disparities driven by both monitoring infrastructure and public data sharing"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5, color = "#2d2d2d"),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "#666666", margin = margin(b = 10)),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm")
  )

# Ecosystem distribution barplot (inset)
p_eco <- ecosystem_data %>%
  filter(ecosystem != "Other") %>%
  count(ecosystem) %>%
  ggplot(aes(x = reorder(ecosystem, n), y = n / 1e6, fill = ecosystem)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  scale_fill_brewer(palette = "Blues", direction = -1) +
  coord_flip() +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = "#d9d9d9", size = 0.5),
    plot.title = element_text(face = "bold", size = 10, color = "#2d2d2d"),
    axis.title = element_text(size = 8, color = "#666666"),
    axis.text = element_text(size = 8, color = "#2d2d2d")
  ) +
  labs(
    title = "Observations by Ecosystem",
    x = NULL,
    y = "Observations (Millions)"
  )

# Place the inset barplot in the Russia/China area (upper right panel coordinates)
p_combined <- p_map + inset_element(p_eco, left = 0.58, bottom = 0.52, right = 0.96, top = 0.88)
print(p_combined)
```

---

# 2. Figure 2: Sampling Objectives & Mismatched Frequencies

### Conceptual Description
* **Aims:** Demonstrate that even within the same geographic region (Wisconsin), sampling designs are fragmented and serve different objectives. This shows the contrast between daily recreational monitoring, monthly regulatory compliance, and quarterly trend analysis.
* **Visual Components:**
  * **Left Panel (Map):** A zoomed-in map of Wisconsin plotting three freshwater sites monitored at contrasting frequencies:
    1. **Swimming/Recreation Site (Daily):** Wisconsin DNR inland lake beach (swimming-season daily sampling).
    2. **Regulatory Attainment Site (Monthly):** Wisconsin DNR river compliance site.
    3. **Long-Term Trend Site (Quarterly):** Wisconsin DNR river trend site.
  * **Right Panel (Timeseries):** Stacked timeseries plots showing the actual observation intervals for these three sites in 2022–2023.
* **Key Findings & Concept:**
  * **Swimming (Daily):** Protecting swimmers from pathogen exposure requires high-frequency (daily) monitoring to capture transient beach closures.
  * **Regulatory Compliance (Monthly):** Checking if a waterbody meets Clean Water Act water quality standards is typically satisfied by monthly regulatory grabs.
  * **Long-Term Trends (Quarterly):** Tracking multi-year effects of climate change or agricultural policy can be satisfied by low-frequency quarterly samples.
  * Tying these frequencies to different management objectives highlights why "abundant data" (which is mostly monthly/quarterly) cannot be used to protect recreational swimmers.

### Executable R Code for Figure 2

```{r figure_2, fig.width=12, fig.height=7}
# Load WI map boundaries
ca_map <- ne_states(country = "United States of America", returnclass = "sf") %>%
  filter(name == "Wisconsin")

# Define three representative freshwater sites in Wisconsin (Wisconsin DNR)
site_coords <- data.frame(
  site_id = c("WIDNR_WQX-683422", "WIDNR_WQX-543001", "WIDNR_WQX-313038"),
  label = c("1) Swimming (Daily Req.)", "2) Regulatory (Monthly Req.)", "3) Long-Term Trend (Quarterly Req.)"),
  lat = c(42.9374, 42.6091, 44.4585),
  lon = c(-88.4797, -89.0706, -87.5561)
)

site_coords_sf <- st_as_sf(site_coords, coords = c("lon", "lat"), crs = 4326)

# Plot zoomed-in Wisconsin map
p_ca_map <- ggplot() +
  geom_sf(data = ca_map, fill = "#fdfdfd", color = "#b0b0b0", size = 0.3) +
  geom_sf(data = site_coords_sf, aes(color = label, shape = label), size = 5) +
  scale_color_manual(
    values = c(
      "1) Swimming (Daily Req.)" = "#e34a33", 
      "2) Regulatory (Monthly Req.)" = "#31a354", 
      "3) Long-Term Trend (Quarterly Req.)" = "#2b8cbe"
    ),
    name = "Monitoring Objective"
  ) +
  scale_shape_manual(
    values = c(
      "1) Swimming (Daily Req.)" = 16, 
      "2) Regulatory (Monthly Req.)" = 17, 
      "3) Long-Term Trend (Quarterly Req.)" = 15
    ),
    name = "Monitoring Objective"
  ) +
  coord_sf(xlim = c(-93.0, -86.5), ylim = c(42.3, 47.2), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.box.background = element_rect(color = "#d9d9d9", size = 0.5),
    panel.background = element_rect(fill = "#e8f2ff", color = NA), # Lake/water fill
    plot.title = element_text(face = "bold", size = 12, color = "#2d2d2d")
  ) +
  labs(
    title = "A) Wisconsin Sites",
    x = "Longitude", y = "Latitude"
  )

# Extract observations for 2022-2023 for these three sites
ts_data <- clean_data %>%
  filter(site_id %in% c("WIDNR_WQX-683422", "WIDNR_WQX-543001", "WIDNR_WQX-313038")) %>%
  collect() %>%
  filter(date >= as.Date("2022-01-01") & date <= as.Date("2023-12-31")) %>%
  mutate(
    label = case_when(
      site_id == "WIDNR_WQX-683422" ~ "1) Swimming (Daily Req.)",
      site_id == "WIDNR_WQX-543001" ~ "2) Regulatory (Monthly Req.)",
      site_id == "WIDNR_WQX-313038" ~ "3) Long-Term Trend (Quarterly Req.)"
    )
  )

# Generate stacked timeseries plot
p_ts <- ggplot(ts_data, aes(x = date, y = val, color = label, shape = label)) +
  geom_line(size = 0.4, alpha = 0.5) +
  geom_point(size = 2) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(
    values = c(
      "1) Swimming (Daily Req.)" = "#e34a33", 
      "2) Regulatory (Monthly Req.)" = "#31a354", 
      "3) Long-Term Trend (Quarterly Req.)" = "#2b8cbe"
    )
  ) +
  scale_shape_manual(
    values = c(
      "1) Swimming (Daily Req.)" = 16, 
      "2) Regulatory (Monthly Req.)" = 17, 
      "3) Long-Term Trend (Quarterly Req.)" = 15
    )
  ) +
  facet_wrap(~label, ncol = 1, scales = "free_y") +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 9, hjust = 0),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12, color = "#2d2d2d")
  ) +
  labs(
    title = "B) Observations (2022-2023)",
    x = "Date",
    y = "E. coli Concentration (CFU/100mL)"
  )

# Combine map and timeseries side-by-side
p_fig2_combined <- p_ca_map + p_ts + plot_layout(widths = c(1, 1.3))
print(p_fig2_combined)
```

---

# 3. Figure 3: Quantifying the Actionability Gap

### Conceptual Description
* **Aims:** Provide a quantitative summary of the data density bottleneck and sampling frequency limitations across the entire global dataset.
* **Visual Components:**
  * **Panel A (Data Volume):** A cumulative distribution function (CDF) or step-survival curve showing the percentage of sites that survive as the minimum observation threshold ($N$) increases from 1 to 1,000.
  * **Panel B (Sampling Interval):** A bar plot of the average sampling return interval (Daily, Weekly, Monthly, Quarterly, Yearly, $> \text{Yearly}$) for all multi-sample sites.
* **Key Findings:**
  * Most sites have very few observations: **only 10.8% of sites have $\ge 100$ observations**, and only **0.7% have $\ge 500$ observations**.
  * Among sites with at least 2 samples, **only 7.0% are monitored on a daily or near-daily basis** ($\le 1.5$ days). The vast majority are monitored **monthly (29.9%)** or **quarterly (30.7%)**, which fails to match the fast-changing kinetics of waterborne pathogen spikes.

### Executable R Code for Figure 3

```{r figure_3, fig.width=12, fig.height=6}
# Group by site and calculate observation counts and spans
site_summary <- clean_data %>%
  select(site_id, lat, lon, date) %>%
  collect() %>%
  mutate(
    site_key = if_else(is.na(site_id), paste0("coord_", lat, "_", lon), site_id)
  ) %>%
  group_by(site_key) %>%
  summarise(
    n_obs = n(),
    min_date = min(date),
    max_date = max(date),
    span_days = as.numeric(max_date - min_date),
    .groups = "drop"
  )

# Part A: Threshold survival curve
thresholds <- c(1, 2, 5, 10, 20, 50, 100, 250, 500, 1000)
survival_df <- data.frame(
  threshold = thresholds,
  sites = sapply(thresholds, function(t) sum(site_summary$n_obs >= t))
) %>%
  mutate(pct = (sites / nrow(site_summary)) * 100)

p_survival <- ggplot(survival_df, aes(x = factor(threshold), y = pct, group = 1)) +
  geom_line(color = "#3182bd", size = 1.1) +
  geom_point(color = "#08519c", size = 3) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12, color = "#2d2d2d"),
    axis.title = element_text(size = 9, color = "#666666"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "A) Site Data Volume Survival Curve",
    x = "Minimum Observations per Site",
    y = "Percentage of Global Sites (%)"
  )

# Part B: Sampling interval distribution
interval_df <- site_summary %>%
  filter(n_obs >= 2) %>%
  mutate(
    avg_interval = span_days / (n_obs - 1)
  ) %>%
  mutate(
    interval_cat = case_when(
      avg_interval <= 1.5 ~ "Daily\n(<=1.5d)",
      avg_interval > 1.5 & avg_interval <= 8 ~ "Weekly\n(1.5-8d)",
      avg_interval > 8 & avg_interval <= 31 ~ "Monthly\n(8-31d)",
      avg_interval > 31 & avg_interval <= 92 ~ "Quarterly\n(31-92d)",
      avg_interval > 92 & avg_interval <= 366 ~ "Yearly\n(92-366d)",
      avg_interval > 366 ~ "Longer\n(>366d)",
      TRUE ~ "Unknown"
    )
  ) %>%
  count(interval_cat) %>%
  mutate(pct = (n / sum(n)) * 100)

# Reorder categories logically
interval_df$interval_cat <- factor(
  interval_df$interval_cat, 
  levels = c("Daily\n(<=1.5d)", "Weekly\n(1.5-8d)", "Monthly\n(8-31d)", "Quarterly\n(31-92d)", "Yearly\n(92-366d)", "Longer\n(>366d)")
)

p_intervals <- ggplot(interval_df, aes(x = interval_cat, y = pct, fill = interval_cat)) +
  geom_col(show.legend = FALSE, width = 0.55) +
  scale_fill_brewer(palette = "Blues", direction = -1) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12, color = "#2d2d2d"),
    axis.title = element_text(size = 9, color = "#666666"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "B) Distribution of Sampling Return Intervals",
    x = "Sampling Interval Category",
    y = "Percentage of Multi-Sample Sites (%)"
  )

# Combine plots side-by-side
p_fig3_combined <- p_survival + p_intervals
print(p_fig3_combined)
```

---

# Figure 4: Reclaiming Recreation Days through High-Frequency Monitoring

### Conceptual Description
* **Aims:** Demonstrate that bacterial pollution in recreational waters is not a permanent state, but rather a transient condition (usually weather- or discharge-driven). High-frequency (daily) data allows water managers to safely "reclaim" recreation days that are otherwise lost to conservative, long-term regulatory closures or public distrust.
* **Visual Components:**
  * **Panel A (Freshwater High-Frequency Site Safety):** A donut chart summarizing the safety status of all samples collected at high-frequency freshwater *E. coli* sites (631 sites, 235,398 samples). It highlights that even at these intensively-monitored sites, the water is safe for contact recreation **77.6% of the time** (concentration $\le 235$ CFU/100mL).
  * **Panel B (Freshwater Case Study):** A timeseries of a Lake Erie swimming beach (Ohio) during the Summer of 2023. It illustrates the transient nature of pollution spikes (exceeding the US EPA freshwater Beach Action Value of 235 CFU/100mL) and shows how quickly water quality recovers to safe levels (typically within ~48 hours).
* **Key Findings & Concept:**
  * Pollution is a passing state, not a permanent attribute of most beaches.
  * With monthly sampling, a single bad sample can impair a beach for an entire month. High-frequency daily sampling allows managers to target closures precisely (≈22% of the time) and keep the waters open for recreation on the remaining ≈78% of safe days.

### Executable R Code for Figure 4

```{r figure_4, fig.width=12, fig.height=6}
# Panel A: Create data for high-frequency freshwater site safety donut chart
# Freshwater E. coli sites with median sampling gap <= 3 d and n >= 30:
# 631 sites, 235,398 samples -> 77.6% safe (<= 235 CFU/100mL), 22.4% unsafe (> 235)
safe_df <- data.frame(
  Status = c("Safe (<235 CFU)", "Unsafe (>235 CFU)"),
  Percentage = c(77.6, 22.4)
)

p_safe <- ggplot(safe_df, aes(x = "", y = Percentage, fill = Status)) +
  geom_col(width = 0.55, color = "white", size = 1) +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = c("Safe (<235 CFU)" = "#31a354", "Unsafe (>235 CFU)" = "#de2d26")) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5, color = "#2d2d2d")
  ) +
  labs(title = "A) High-Frequency Freshwater E. coli Safety\n(631 sites, 235,398 samples)")

# Panel B: Case study timeseries for Lake Erie beach 21OHBCH-OHEUSP01 (Summer 2023)
beach_ts <- clean_data %>%
  filter(site_id == "21OHBCH-OHEUSP01") %>%
  collect() %>%
  filter(lubridate::year(date) == 2023 & lubridate::month(date) %in% 5:9) %>%
  group_by(date) %>%
  summarise(val = max(val), .groups = "drop")

p_beach <- ggplot(beach_ts, aes(x = date, y = val)) +
  geom_hline(yintercept = 235, linetype = "dashed", color = "#de2d26", size = 0.8) +
  geom_line(color = "gray60", size = 0.6) +
  geom_point(aes(color = val > 235), size = 2.5) +
  annotate("text", x = as.Date("2023-05-05"), y = 300, label = "EPA Beach Action Value (235)",
           color = "#de2d26", fontface = "italic", size = 3.5, hjust = 0) +
  scale_color_manual(values = c("FALSE" = "#31a354", "TRUE" = "#de2d26")) +
  scale_y_log10(labels = scales::comma) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 12, color = "#2d2d2d"),
    axis.title = element_text(size = 9, color = "#666666"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "B) Case Study: Lake Erie Beach, Ohio (Summer 2023)",
    subtitle = "High-frequency monitoring reveals transient spikes; water recovers to safe levels within ~48 hours",
    x = "Date",
    y = "E. coli (CFU/100mL) [Log Scale]"
  )

# Combine plots side-by-side
p_fig4_combined <- p_safe + p_beach + plot_layout(widths = c(1, 1.8))
print(p_fig4_combined)
```

---

# 5. Discussion & Synthesis Plan for the Paper

### 1. The Public Data Access and Sharing Gap
* The apparent "data sparsity" in regions like Russia, China, and parts of South America is likely due to **portal accessibility barriers** rather than a true absence of monitoring. These countries maintain major industrial, agricultural, and urban water centers, which require active monitoring. However, their data is locked in national agencies or offline databases. 
* *Scientific Recommendation:* The paper should argue that resolving global water security requires standardized international data-sharing treaties, similar to global meteorological reporting (WMO), rather than just funding new monitoring stations.

### 2. The Public Health Mismatch (Swimming vs. Standards)
* Regulatory thresholds (such as the EPA limit of 235 CFU/100mL for E. coli) are based on gastrointestinal illness risks. However, applying these limits to weekly or monthly samples creates a **high rate of false negatives** (missing transient pathogen spikes) and **false positives** (keeping beaches closed long after storm runoff has cleared).
* Because standard incubation-based laboratory tests require 24 hours to return results, beach closures are always **reactive** (representing yesterday's water quality).

### 3. Policy and Future Research Recommendations
* **Event-Based Sampling:** Move away from rigid weekly calendars to weather-driven/precipitation-triggered sampling.
* **Predictive Modeling:** Develop machine learning/statistical models using rainfall, turbidity, and river discharge to predict water safety in real-time.
* **Rapid Diagnostic Integration:** Push for the widespread adoption of qPCR and automated enzymatic sensors to replace legacy 24-hour incubation methods.
